Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
Minor edit

\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!75} RNA\textcolor{blue!70!black}{Blue}Print}}
\newcommand{\ourprog}{{\tt \bfseries{}\color{black!75}RNA\textcolor{red!70!black}{Red}Print}}

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling.  A) Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for sampled sequences; the good
    match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures S1 and S2 of subfigure B.  B) D\&P energy
    distributions for two target structures S1 and S2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures (annotated in red, yellow, and blue
    resepectively)---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations. As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design. 
As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}
where $\EnsE(S)$, the \emph{ensemble free energy} of $S$, is computed
by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
pseudoknotted targets, $\EnsE(S)$ is approximated by the \emph{minimum
  free energy} of S as estimated by {\tt HotKnots} in the energy model
of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
captures the distance of the targets from the ensemble free energy,
while the second term penalizes the dispersion of targets; \Obj{} is
best (minimal) when all targets simultaneously achieve the minimum
free energy of the sequence.

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, confirm that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
their MFE, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

@@

\begin{figure}
  {\includegraphics[width=\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
  \caption{Comparison of the \Obj{} (see
    Eq.~\eqref{eq:blueprintobjective}; smaller values $\to$ better
    designs) for sequences sampled by \ourprog{} and uniform sampling
    (\RNAblueprint) for six benchmark sets. For both sampling schemes,
    we show the MultiDefects of the sampled sequences, as well as
    results after further optimization by local search (``+ Opt.'').}
\label{fig:benchmark-results}
\end{figure}

\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
computing the number of valid sequences for a given compatibility
graph $G=(V,E)$. Note that this problem corresponds to the partition
function problem in a simple base pair model, assigning $0$ to
canonical and $\infty$ to non-canonical base pairs and setting 
$\beta>0$, so that any valid structure counts as $1$ in the partition 
function. As previously noted~\citep{Flamm2001}, a set of target
structures admits a valid design iff its compatibility graph is
bipartite, which can be tested in linear time.  Moreover, without
$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
of $G$ would be entirely determined by the assignment of a single of
its nucleotides. The number of valid designs is thus simply
$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
  connected components}.

>>>>

\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!85} RNA\textcolor{blue!70!black}{Blue}Print}}
\newcommand{\ourprog}{{\tt \bfseries{}\color{black!85}RNA\textcolor{red!70!black}{Red}Print}}

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically  ($\gamma=1.2$) to achieve effective optimization.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its good empirical behavior.

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling.  A) Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for sampled sequences; the good
    match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures S1 and S2 of subfigure B.  B) D\&P energy
    distributions for two target structures S1 and S2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures (annotated in red, yellow, and blue
    resepectively)---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations. As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design. 
As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    \Obj(S) &= \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}
where $\EnsE(S)$, the \Def{ensemble free energy} of $S$, is computed
by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
pseudoknotted targets, $\EnsE(S)$ is approximated by the \Def{minimum
  free energy} (MFE) of S as estimated by {\tt HotKnots} in the energy model
of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
captures the distance of the targets from the ensemble free energy,
while the second term penalizes the dispersion of targets; \Obj{} is
best (minimal) when all targets simultaneously achieve the minimum
free energy of the sequence.

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

@@

\begin{figure}
  {\includegraphics[width=.9\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
  \caption{Comparison of the \Obj{} (see
    Eq.~\eqref{eq:blueprintobjective}; smaller values $\to$ better
    designs) for sequences sampled by \ourprog{} and uniform sampling
    (\RNAblueprint) for six benchmark sets. For both sampling schemes,
    we show the MultiDefects of the sampled sequences, as well as
    results after further optimization by local search (``+ Opt.'').}
\label{fig:benchmark-results}
\end{figure}

\section{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
computing the number of valid sequences for a given compatibility
graph $G=(V,E)$. Note that this problem corresponds to the partition
function problem in a simple base pair model, assigning $0$ to
canonical and $\infty$ to non-canonical base pairs and setting 
$\beta>0$, so that any valid structure counts as $1$ in the partition 
function. As previously noted~\citep{Flamm2001}, a set of target
structures admits a valid design iff its compatibility graph is
bipartite, which can be tested in linear time.  Moreover, without
$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
of $G$ would be entirely determined by the assignment of a single of
its nucleotides. The number of valid designs is thus simply
$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
  connected components}.
